Analyzing Trends in Crime Against Persons: Insights from the Toronto Police Annual Statistical Report

### 2024-01-25 outputs/paper/paper.qmd
Final Draft

---
title: "Analyzing Trends in Crime Against Persons: Insights from the Toronto Police Annual Statistical Report"
author: "Mohammed Yusuf Shaikh"
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "Public Safety and security are a paramount concern in the 20th century, the analysis of statistics in Victimology helps to visualize and derive the conclusions of urban crime dynamics. This analytical report delves into the trends of victims of a certain crime in Toronto from the year 2014 to 2022. Through a series of data visualization approach to examine the relationship between crime incidents and demographic factors. The data reveals significant fluctuations in crime rates in correspond to the socio-economic shifts, including a notable dip in crimes reported during pandemic between 2019-2020. The visualization of the certain trends provides a compelling narrative on the dynamics of victims of crime in Toronto, advocating for all the policymakers to call for a reevaluation of current approaches to public safety and crime prevention."
format: pdf
fig-pos: "H"
number-sections: true
bibliography: references.bib
---

@@

# Introduction
The concern of Public safety in the Urban society is a necessary step towards safeguarding people of that community. The city of Toronto has various socio-economic challenges one of them being Crime against a person. A crime incident might leave a victim and our report aims to explore the patterns and trend of crimes against the person. The paper aims to understand how this trends have evolved among victims and analyzing there  demograpics from 2014 to 2022. To elaborate, the report context is not only quantifying crime victims but also interpret and explain whereby it motivates to a better approach towards urban safety and policies. The Criminal victimization is a frightening and unsettling experience for many Canadians (Cite).

Victimology is a vast sub-category of criminology to understand various emotional factors and its causation, however based on our police annual report – victims of  crime showcase a statistical record based on which the report will share the knowledge to work on the gap to understand between victim demographics and crime. This report addresses this gap by visualizing crime statistics through a demographic lens, thereby unveiling patterns that may not be evident in numerical data alone

Through R programming language [@R] we analyze the data obtained from Open data Toronto Portal [@rOpenDataToronto]. This dataset provides a detailed account of all recorded count of victim under the jurisdiction of the Toronto Police Services over an eight-period. Our research offers an examination of cleaned data points and produced visual formats that reveal the shapes the crime type as well as their victims across different genders and age cohorts. The analysis, supplemented with various analytical tools like Tidyverse [@rTidyverse], AER [@rAER], Janitor [@rJanitor], and Knitr [@rKnitr], delves deep into the trends, revealing patterns that could have implications for policy formulation and implementation.

The Introduction is followed by the Data section. Within the data section we talk about the data source used for this report, the data cleaning process applied and the result that is the visual analysis of trends and pattern involved. The Discussion section shares additional insights into the data findings, as well as discussing measurement errors and limitations of this analysis. Lastly, the document summarizes the analysis.

@@

The data used in this paper was collected by the OpenDataToronto Library [@rOpenDataToronto]. The specific dataset used in this research is the
'Police-Annual-Statistical-Report-Victims-of-Crimes' [@citeDataset] from the year 2014 to 2022. The dataset is categorized by crime against a person, so subtype includes: Assault, Robbery, Sexual Violation, and Other offenses. The stratification of the dataset follows victim demographics such as age group, age cohort and gender.

@@

The Table 1 showcases the cleaned dataset from the raw data extracted from the 'Police-Annual-Statistical-Report-Victims-of-Crimes' [@citeDataset].

@@

# Use kable from knitr to create a simple table

@@

# Creating the line graph
ggplot(summarized_data, aes(x = report_year, y = Total_Count)) +
  geom_line(color = "blue") +  # This creates the line graph
  geom_point() +  # Optional: adds points at each year
  labs(title = "Total Victims of Crime",
       x = "Year",
       y = "Total Victims of Crime") +
  theme_minimal()
```

@@

### By Gender
@fig-victims-by-Gender displays the distribution of victims of crimes grouped by gender across different years, from 2014 to 2022. The bars represent the total count of victims for each gender category, with separate colors for females (F), males (M), and an Unknown category (U). We can infer for every year from 2014-22 Female victims have higher count compared to Male victims. The data also showcase significantly lower number of victims as the remain ‘Unknown’.  The figure high variation in terms of victims of crime over year. There isn’t any clear pattern in the data.A notable decrease in victim count for all genders can be seen around 2020. This could be an area of further investigation, as external factors such as the COVID-19 pandemic may have influenced crime rates and reporting.

@@

```{r}
#| label: tbl-Avg-Victim-of-crime-by-age-group
#| tbl-cap: Average Victims by Age Group (2014-2022)
#| echo: false

cleaned_dataset |>
  group_by(age_group) |>
  summarize(
    mean_Victim_count = mean(victim_count),
  ) |>
  kable(
    col.names = c("Crime", "Average Victim Count"),
    booktabs = TRUE
    )

@@

# Assuming cleaned_dataset is your data frame and it has the columns 'report_year', 'age_group', and 'victim_count'
average_victims_per_year_age_group <- cleaned_dataset %>%
  group_by(report_year, age_group) %>%
  summarize(avg_victim_count = mean(victim_count)) %>%
  ungroup()

# To visualize this data, you can create a line plot for each age group
ggplot(average_victims_per_year_age_group, aes(x = report_year, y = avg_victim_count, group = age_group, color = age_group)) +
  geom_line() +  # Add lines for each age group
  geom_point() +  # Add points
  labs(title = "Average Number of Victims per Year by Age Group",
       x = "Year",
       y = "Average Number of Victims",
       color = "Age Group") +
  theme_minimal() +
  theme(legend.position = "bottom")  # Position the legend at the bottom

@@

# Summarize the total count by age cohort and subtype
crimes_age_subtype <- cleaned_dataset %>%
  group_by(age_cohort, subtype) %>%
  summarise(total_count = sum(victim_count)) # Ensure grouped data is ungrouped after summarization

# Create a horizontal bar graph for Crimes by Age Cohort and Subtype
ggplot(crimes_age_subtype, aes(x = total_count, y = reorder(age_cohort, total_count), fill = subtype)) +
  geom_bar(stat = "identity") +
  labs(title = "Victim of Crime by Age Cohort and Crime Subtype",
       x = "Total Count",
       y = "Age Cohort") +
  theme_minimal() +
  theme(legend.position = "bottom") # Adjust legend position if necessary

>>>>

---
title: "Analyzing Trends in Crime Against Persons: Insights from the Toronto Police Annual Statistical Report"
author: "Mohammed Yusuf Shaikh"
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "Public Safety and security are a paramount concern in the 21st century, the analysis of statistics in Victimology helps to visualize and derive the conclusions of urban crime dynamics. This analytical report delves into the trends of victims of a crime in Toronto from the year 2014 to 2022. This report employs a series of data visualization approaches to examine the relationship between crime incidents and demographic factors. The data reveals significant fluctuations in crime rates in correspond to the socio-economic shifts, including a notable dip in crimes reported during pandemic between 2019-2020. The visualization of the certain trends provides a compelling narrative on the dynamics of victims of crime in Toronto, advocating for all the policymakers to call for a reevaluation of current approaches to public safety and crime prevention."
format: pdf
fig-pos: "H"
toc: true
number-sections: true
bibliography: references.bib
---

@@

# Introduction

Concern of public safety in the urban societies is a necessary step towards safeguarding the residents of that community. The city of Toronto has various socio-economic challenges one of them being crime against a person. A crime incident might leave a victim and our report aims to explore the patterns and trend of crimes against the person. The paper aims to understand how this trends have evolved among victims and to analyze their demographics from 2014 to 2022. Furthermore, it's important to note that the report's scope extends beyond merely quantifying crime victims but it also interpret and explain whereby it motivates to a better approach towards urban safety and policies. The Criminal victimization is a frightening and unsettling experience for many Canadians (@statscanada).
Victimology is a vast sub-category of criminology to understand various emotional factors and its causation, however based on our police annual report – victims of  crime showcase a statistical record based on which the report will share the knowledge to work on the gap to understand between victim demographics and crime. This report addresses this gap by visualizing crime statistics through a demographic lens, thereby unveiling patterns that may not be evident in numerical data alone. Through R programming language [@R], in this report, we analyze the data obtained. from Open data Toronto Portal [@rOpenDataToronto]. This dataset provides a detailed account of all recorded count of victim under the jurisdiction of the Toronto Police Services over an eight-period. Our research offers an examination of cleaned data points and produced visual formats that reveal the shapes the crime type as well as their victims across different genders and age cohorts. The analysis, supplemented with various analytical tools like Tidyverse [@rTidyverse], AER [@rAER], Janitor [@rJanitor], and Knitr [@rKnitr], delves deep into the trends, revealing patterns that could have implications for policy formulation and implementation. The introduction is followed by the data section. Within the data section we talk about the data source used for this report, the data cleaning process applied and the result that is the visual analysis of trends and pattern involved. The Discussion section shares additional insights into the data findings, as well as discussing measurement errors and limitations of this analysis. Lastly, the document summarizes the analysis with conclusion.

@@

The data used in this paper were collected by the OpenDataToronto Library [@rOpenDataToronto]. The specific dataset used in this research is the
'Police-Annual-Statistical-Report-Victims-of-Crimes' [@citeDataset] from the year 2014 to 2022. The dataset is categorized by crime against a person, so subtype includes: assault, robbery, sexual Violation, and other offenses. The stratification of the dataset follows victim demographics such as age group, age cohort, and gender.

@@

The Table 1 (@tbl-Cleaned-data-output) showcases the cleaned dataset from the raw data extracted from the 'Police-Annual-Statistical-Report-Victims-of-Crimes' [@citeDataset].

@@

# Making table from Kable for pdf paper

@@

# Creating the line graph showing of the Year & Total Victim over 8 year period
ggplot(summarized_data, aes(x = report_year, y = Total_Count)) +
  geom_line(color = "blue") +  # This creates the line graph
  geom_point() +  # Optional: adds points at each year
  labs(title = "Total Victims of Crime",
       x = "Year",
       y = "Total Victims of Crime") +
  theme_minimal()
```

@@

### By Gender
@fig-victims-by-Gender displays the distribution of victims of crimes grouped by gender across different years, from 2014 to 2022. The bars represent the total count of victims for each gender category, with separate colors for females (F), males (M), and an Unknown category (U). We can infer for every year from 2014-22 Female victims have higher count compared to Male victims. The data also showcase significantly lower number of victims as the remain ‘Unknown’.  The figure high variation in terms of victims of crime over year. There isn’t any clear pattern in the data.A notable decrease in victim count for all genders can be seen around 2020. This could be an area of further investigation, as external factors such as the COVID-19 pandemic may have influenced crime rates and reports conducted during that year.

@@

```{r}
#| label: tbl-Avg-Victim-of-crime-by-age-group
#| tbl-cap: Average Victims by Age Group (2014-2022)
#| echo: false
# Using Cleaned Data set and grouping vaiable of interest
cleaned_dataset |>
  group_by(age_group) |>
  summarize(
    mean_Victim_count = mean(victim_count),
  ) |>
  kable(
    col.names = c("Crime", "Average Victim Count"),
    booktabs = TRUE
    )

@@

# Using cleaned_dataset that has the columns 'report_year', 'age_group', and 'victim_count'
average_victims_per_year_age_group <- cleaned_dataset %>%
  group_by(report_year, age_group) %>%
  summarize(avg_victim_count = mean(victim_count)) %>%
  ungroup()

# To visualize - a line graph for each age group
ggplot(average_victims_per_year_age_group, aes(x = report_year, y = avg_victim_count, group = age_group, color = age_group)) +
  geom_line() +  
  geom_point() +  
  labs(title = "Average Number of Victims per Year by Age Group",
       x = "Year",
       y = "Average Number of Victims",
       color = "Age Group") +
  theme_minimal() +
  theme(legend.position = "bottom")

@@

# Making subtables for graph
crimes_age_subtype <- cleaned_dataset %>%
  group_by(age_cohort, subtype) %>%
  summarise(total_count = sum(victim_count))

# Horizontal bar graph for Crimes and its victim Age Cohort and Subtype
ggplot(crimes_age_subtype, aes(x = total_count, y = reorder(age_cohort, total_count), fill = subtype)) +
  geom_bar(stat = "identity") +
  labs(title = "Victim of Crime by Age Cohort and Crime Subtype",
       x = "Total Count",
       y = "Age Cohort") +
  theme_minimal() +
  theme(legend.position = "bottom")
